---
title: "earthquakevisuals-vignette"
output: rmarkdown::html_vignette
fig_width: 20
fig_height: 13
vignette: >
  %\VignetteIndexEntry{earthquakevisuals-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Setup

After installing the package the required packages must be loaded.

```{r setup}
library(earthquakevisuals)
library(ggplot2)
```

The data is stored under the name `noaa_earthquakes`. It contains observations from 2150 B.C. up until 2025. To be able to use it properly some preprocessing steps are required bevorehand. For this purpose the two functions `eq_clean_data()` and `eq_location_clean()` can be used. Among others the former creates a date column and the latter separates the country name and the location name into distinct columns.

```{r preprocessing}
data <- noaa_earthquakes |> 
    eq_clean_data() |>
    eq_location_clean()
```


## Ggplot2 dotplots

Once this is done the first visualization can be created. The package contains the custom ggplot2 geom `geom_timeline()` which creates a horizontal dotplot. It also provides the possibility to filter the data in the geom itself via the parameters `xmin` and `xmax`.

```{r geom_timeline}
data |> 
    subset(country %in% c("ALASKA", "CHINA", "JAPAN")) |> 
    ggplot(aes(date)) +
    geom_timeline(xmin = as.Date("2000-01-01"), 
                  xmax = as.Date("2018-01-01")
                  ) +
    theme_classic() +
    theme(axis.line.y = element_blank(), 
          axis.ticks.y = element_blank(), 
          axis.text.y = element_blank()
          ) +
    labs(x = "DATE", 
         y = element_blank(), 
         title = "NOAA Earthquakes") +
    ylim(0.5, 2)
```

But `geom_timeline()` allows for much more customization. E. g. you can give a `y aesthetic` to split the dotplots into different groups. You can also specify size and color to include more dimensions.
Lastly, the `geom_timeline_label()` function can be used to add a label to each earthquake. Via the parameter `n_max` only the biggest earthquakes are labelled according to the `size aesthetic`. 

```{r geom_timeline_label}
data |> 
    subset(country %in% c("ALASKA", "CHINA", "JAPAN")) |> 
    ggplot(aes(date, country, size = mag, color = total_deaths)) +
    geom_timeline(xmin = as.Date("2000-01-01"), 
                  xmax = as.Date("2018-01-01")
                  ) +
    geom_timeline_label(aes(label = location_name),
                        xmin = as.Date("2000-01-01"), 
                        xmax = as.Date("2018-01-01"),
                        n_max = 5
                        ) +
    theme_classic() +
    theme(axis.line.y = element_blank(),
          axis.ticks.y = element_blank(),
          legend.position = "bottom"
          ) +
    scale_size_continuous(name = "Richter scale value") +
    scale_color_continuous(name = "# deaths") +
    labs(x = "DATE", 
         y = element_blank(), 
         title = "NOAA Earthquakes")
```


## Leaflet maps

The function `eq_map()` is useful when one wants to display the earthquakes on a map. The parameter `annot_col` steers which column is used for labelling the individual earthquakes.

```{r leaflet_1}
data |> 
    subset(country == "INDIA" & date >= as.Date("2000-01-01")) |>
    eq_map(annot_col = "date")
```

To have a label with more information the function `eq_create_label()` can be used to generate a HTML formatted column with information about an earthquake's location, its magnitude and the associated number of deaths. The newly created column can serve as an input for the `eq_map()` function.

```{r}
data |> 
    transform(popup_text = eq_create_label(data)) |> 
    subset(country == "INDIA" & date >= as.Date("2000-01-01")) |> 
    eq_map(annot_col = "popup_text")
```
